---
title: "DeadWood_snagDecay"
author: ""
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DeadWood_snagDecay}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Overview

`DeadWood_snagDecay` advances standing dead trees (snags) through five decay
classes (DC1–DC5) on an annual timestep using a species-specific Markov
transition matrix. At each timestep, snags that fall are transferred to
`sim$fallenSnags` for ingestion by `DeadWood_DWDDecay`.

## Inputs

| Object | Class | Description |
|--------|-------|-------------|
| `cohortData` | `data.table` | Pixel-level mortality table: `pixelID`, `year`, `species`, `B` (Mg/ha) |

## Outputs

| Object | Class | Description |
|--------|-------|-------------|
| `snagTable` | `data.table` | Current snag inventory: `pixelID`, `species`, `DC`, `ageInDC`, `initBiomass` |
| `fallenSnags` | `data.table` | Snags that fell this timestep — same schema as `snagTable` |

## Parameters

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `snagTransMat` | `matrix` | 5×5 zero matrix | Annual DC transition probability matrix |
| `snagFallProb` | `numeric` | `rep(0.1, 5)` | Annual fall probability by DC (length 5) |
| `species` | `character` | `"Pinus strobus"` | Species to filter from `cohortData` |

## Example

```{r example}
library(SpaDES.core)
library(data.table)

cohortData <- data.table(
  pixelID = 1L, year = 1L, species = "Pinus strobus", B = 20.0
)

mySim <- simInit(
  times   = list(start = 0, end = 10),
  modules = list("DeadWood_snagDecay"),
  params  = list(DeadWood_snagDecay = list(
    snagTransMat = myTransMat,   # supply a real 5x5 matrix
    snagFallProb = c(0.09, 0.09, 0.10, 0.12, 0.30),
    species      = "Pinus strobus"
  )),
  objects = list(cohortData = cohortData),
  paths   = list(modulePath = "path/to/modules")
)

mySim <- spades(mySim)
```
