---
title: "DeadWood_snagDecay"
author: ""
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DeadWood_snagDecay}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Overview

`DeadWood_snagDecay` advances standing dead trees (snags) through five decay
classes (DC1–DC5) every 5 years using species-specific Markov transition
matrices. At each timestep, snags that fall are stochastically selected by
decay class and transferred to `sim$fallenSnags` for ingestion by
`DeadWood_DWDDecay`.

## Inputs

| Object | Class | Description |
|--------|-------|-------------|
| `cohortData` | `data.table` | Pixel-level mortality table: `pixelID`, `year`, `species`, `biomass` (Mg/ha), `diameter_cm` (cm) |

## Outputs

| Object | Class | Description |
|--------|-------|-------------|
| `snagTable` | `data.table` | Current snag inventory: `pixelID`, `species`, `DC`, `ageInDC`, `initBiomass`, `diameter_cm` |
| `fallenSnags` | `data.table` | Snags that fell this timestep — same schema as `snagTable` |

## Parameters

| Parameter | Type | Default source |
|-----------|------|----------------|
| `snagTransMat` | `list` | Vanderwel et al. 2006, Table 3 |
| `snagFallProb` | `list` | Vanderwel et al. 2006, Table 2 |
| `species` | `character` | — |
| `defaultDiameter_cm` | `numeric` | — |

### `snagTransMat`

A named list of 5×5 Markov transition matrices, one per species. Each matrix
gives the 5-year conditional probability of moving from one decay class (row)
to another (column). DC5 is absorbing — snags that reach DC5 remain there
until they fall. Row names are `from_DC1`–`from_DC5`; column names are
`to_DC1`–`to_DC5`. Names in the list must match the `species` parameter.

Transitions are strictly forward (lower triangle is zero); a snag cannot
revert to an earlier decay class.

Default matrices (Vanderwel et al. 2006, Table 3):

***Pinus strobus***

| | DC1 | DC2 | DC3 | DC4 | DC5 |
|---|---:|---:|---:|---:|---:|
| **from DC1** | 0.109 | 0.368 | 0.471 | 0.035 | 0.017 |
| **from DC2** | 0.000 | 0.348 | 0.601 | 0.051 | 0.000 |
| **from DC3** | 0.000 | 0.000 | 0.704 | 0.270 | 0.027 |
| **from DC4** | 0.000 | 0.000 | 0.000 | 0.913 | 0.087 |
| **from DC5** | 0.000 | 0.000 | 0.000 | 0.000 | 1.000 |

***Pinus resinosa***

| | DC1 | DC2 | DC3 | DC4 | DC5 |
|---|---:|---:|---:|---:|---:|
| **from DC1** | 0.109 | 0.368 | 0.471 | 0.035 | 0.017 |
| **from DC2** | 0.000 | 0.234 | 0.434 | 0.252 | 0.081 |
| **from DC3** | 0.000 | 0.000 | 0.704 | 0.270 | 0.027 |
| **from DC4** | 0.000 | 0.000 | 0.000 | 0.913 | 0.087 |
| **from DC5** | 0.000 | 0.000 | 0.000 | 0.000 | 1.000 |

The two species share identical DC1, DC3, DC4, and DC5 rows; they differ at
DC2, where *P. resinosa* transitions more rapidly to DC4–DC5.

### `snagFallProb`

A named list of 5-year fall probability vectors, one per species. Each vector
has length 5, with one probability per decay class. At each 5-year timestep
each snag is independently sampled using its current DC's fall probability;
snags that fall are removed from `snagTable` and added to `fallenSnags`.

Default values (Vanderwel et al. 2006, Table 2):

| Species | DC1 | DC2 | DC3 | DC4 | DC5 |
|---------|---:|---:|---:|---:|---:|
| *Pinus strobus* | 0.130 | 0.164 | 0.212 | 0.325 | 0.200 |
| *Pinus resinosa* | 0.000 | 0.039 | 0.094 | 0.224 | 0.080 |

*P. resinosa* DC1 has a fall probability of zero — freshly dead red pine snags
do not fall immediately. Fall probability generally increases with decay class,
peaking at DC4 for both species.

### `species`

Character vector of species names to process from `cohortData`. Default:
`c("Pinus strobus", "Pinus resinosa")`. Each entry must have a corresponding
named entry in both `snagTransMat` and `snagFallProb`; the module will stop
with an informative error if a name is missing.

### `defaultDiameter_cm`

Fallback piece diameter (cm) applied to all incoming snags when `cohortData`
does not include a `diameter_cm` column. Default: `19.0` cm. Minimum allowed
value is `7.5` cm (the minimum coarse woody debris threshold). Diameter is
passed to `DeadWood_DWDDecay` and controls the logistic transition rate once
a piece falls.

## Example

```{r example}
library(SpaDES.project)
set.seed(42)

# Three mortality pulses at different years fill the full DC spectrum by year 50:
#   Year  1 cohort: 50 years of decay → concentrated in DC4-5
#   Year 15 cohort: 35 years          → mixed DC2-4
#   Year 30 cohort: 20 years          → mostly DC1-2
cohortData <- data.table::data.table(
  pixelID     = 1:60,
  year        = c(rep(1L, 20), rep(15L, 20), rep(30L, 20)),
  species     = rep(c(rep("Pinus strobus", 10), rep("Pinus resinosa", 10)), 3),
  biomass     = c(rep(25.0, 20), rep(20.0, 20), rep(15.0, 20)),
  diameter_cm = c(rep(32.0, 20), rep(26.0, 20), rep(20.0, 20))
)

out <- SpaDES.project::setupProject(
  paths      = list(modulePath = "modules"),
  modules    = "BosunForestEcology/DeadWood_snagDecay@main",
  times      = list(start = 0, end = 50),
  cohortData = cohortData
)

mySim <- SpaDES.core::simInitAndSpades2(out)

library(ggplot2)

dcSummary <- mySim$snagTable[, .(biomass = sum(initBiomass)), by = .(DC, species)][order(species, DC)]

ggplot(dcSummary, aes(x = DC, y = biomass, colour = species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = 1:5) +
  scale_colour_manual(values = c("Pinus strobus" = "darkgreen", "Pinus resinosa" = "saddlebrown")) +
  labs(x = "Decay class", y = "Total snag biomass (Mg/ha)",
       title = "Snag biomass by decay class at year 50", colour = NULL) +
  theme_bw()
```
