---
title: "DeadWood_snagDecay"
author: ""
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# DeadWood_snagDecay

A [SpaDES](https://spades.predictiveecology.org/) module that advances standing
dead trees (snags) through five decay classes (DC1–DC5) using a stochastic
Markov chain model. At each 5-year timestep, snags may progress to a more
advanced decay class and are subject to stochastic falling. Fallen snags are
passed to the `DeadWood_DWDDecay` module as downed woody debris (DWD).

Default parameters are calibrated for *Pinus strobus* (Eastern White Pine) using
Vanderwel et al. (2006).

---

## Inputs

| Object | Class | Description |
|--------|-------|-------------|
| `cohortData` | `data.table` | Pixel-level tree mortality table. Required columns: `pixelID` (integer), `year` (numeric — year of death), `species` (character), `biomass` (numeric, Mg ha⁻¹ at death). Optional column: `diameter_cm` (numeric, cm; see `defaultDiameter_cm` parameter if absent). |

**Minimum diameter:** Any snag with `diameter_cm < 7.5 cm` will cause an error.
This threshold reflects the minimum piece size tracked in Vanderwel et al. (2006).

---

## Outputs

| Object | Class | Description |
|--------|-------|-------------|
| `snagTable` | `data.table` | Current standing snag inventory, updated every 5 years. |
| `fallenSnags` | `data.table` | Snags that fell during the current timestep, consumed by `DeadWood_DWDDecay`. Reset each timestep. |

Both tables share the same schema:

| Column | Type | Description |
|--------|------|-------------|
| `pixelID` | integer | Raster pixel of origin |
| `species` | character | Tree species |
| `DC` | integer | Current decay class (1–5) |
| `ageInDC` | integer | Years spent in the current decay class |
| `initBiomass` | numeric | Biomass at time of death (Mg ha⁻¹); never modified after entry |
| `diameter_cm` | numeric | Piece diameter (cm); never modified after entry |

---

## Parameters

| Parameter | Type | Default | Range | Description |
|-----------|------|---------|-------|-------------|
| `snagTransMat` | matrix | See below | — | 5×5 Markov transition probability matrix for decay class advancement (5-year intervals). Source: Vanderwel et al. (2006) Table 3. |
| `snagFallProb` | numeric | See below | [0, 1] | Length-5 vector of 5-year fall probabilities, one per decay class. Source: Vanderwel et al. (2006) Table 2. |
| `species` | character | `"Pinus strobus"` | — | Species name used to filter rows from `cohortData`. |
| `defaultDiameter_cm` | numeric | `19.0` | ≥ 7.5 | Fallback diameter (cm) assigned to all incoming snags when `cohortData` lacks a `diameter_cm` column. |

### Default transition matrix (`snagTransMat`)

Rows are current DC; columns are DC entered at the next 5-year step (conditional
on the snag not having fallen):

```{r snagTransMat}
matrix(c(
  0.109, 0.368, 0.471, 0.035, 0.017,  # from DC1
  0.000, 0.348, 0.601, 0.051, 0.000,  # from DC2
  0.000, 0.000, 0.704, 0.270, 0.027,  # from DC3
  0.000, 0.000, 0.000, 0.913, 0.087,  # from DC4
  0.000, 0.000, 0.000, 0.000, 1.000   # from DC5
), nrow = 5, byrow = TRUE,
dimnames = list(paste0("from_DC", 1:5), paste0("to_DC", 1:5)))
```

Row sums equal 1.0. DC5 is an absorbing state (snags do not decay beyond DC5;
they can only fall).

### Default fall probabilities (`snagFallProb`)

5-year probability of a snag falling, by decay class:

| DC1 | DC2 | DC3 | DC4 | DC5 |
|-----|-----|-----|-----|-----|
| 0.130 | 0.164 | 0.212 | 0.325 | 0.200 |

---

## Events

The module fires two event types on a 5-year schedule.

### `init` (once, at simulation start)

- Validates that `snagTransMat` is non-zero and `snagFallProb` has length 5.
- Creates empty `snagTable` and `fallenSnags` with the correct schema.
- Schedules the first `transition` event at `start(sim) + 5` with priority 1.

### `transition` (every 5 years, priority 1)

Executes three steps in order:

**Step 1 — Absorb new mortality**

Rows in `cohortData` matching the target `species` whose `year` falls in the
interval `(time - 5, time]` are treated as trees that died during the preceding
5-year period. Each is added to `snagTable` as a DC1 snag with `ageInDC = 0`.

**Step 2 — Advance decay class**

Each snag's decay class is updated by drawing from the row of `snagTransMat`
corresponding to its current DC:

$$\text{DC}_\text{new} \sim \text{Categorical}\!\left(\text{snagTransMat}[\text{DC}_\text{old},\, \cdot\,]\right)$$

Because transition probabilities only move forward (upper-triangular matrix), DC
never decreases. Time in decay class is tracked:

$$\text{ageInDC}_\text{new} = \begin{cases} \text{ageInDC} + 5 & \text{if } \text{DC}_\text{new} = \text{DC}_\text{old} \\ 0 & \text{if } \text{DC}_\text{new} > \text{DC}_\text{old} \end{cases}$$

**Step 3 — Stochastic falling**

Using the **pre-transition** DC, each snag independently draws a fall outcome:

$$\text{Fall} \sim \text{Bernoulli}\!\left(\text{snagFallProb}[\text{DC}_\text{old}]\right)$$

Snags that fall are moved to `fallenSnags`. Snags that do not fall remain in
`snagTable` with their updated DC and `ageInDC`.

---

## Event scheduling and module interactions

This module is designed to run alongside `DeadWood_DWDDecay` and
`DeadWood_Biomass`. Event priorities within each 5-year timestep are:

| Priority | Module | Event | Purpose |
|----------|--------|-------|---------|
| 1 | `DeadWood_snagDecay` | `transition` | Advance DC, produce `fallenSnags` |
| 2 | `DeadWood_DWDDecay` | `receive` | Accept `fallenSnags` into DWD pool |
| 3 | `DeadWood_DWDDecay` | `transition` | Advance DWD DC |
| 4 | `DeadWood_Biomass` | `transition` | Compute biomass rasters |

This ordering guarantees that `fallenSnags` is populated before `DeadWood_DWDDecay`
reads it, and that biomass is computed from the fully updated inventories.

---

## Usage example

```{r usage}
library(SpaDES.core)

# Minimal cohort table
cohortData <- data.table::data.table(
  pixelID    = c(1L, 1L, 2L),
  year       = c(5,  5,  10),
  species    = "Pinus strobus",
  biomass    = c(12.5, 8.3, 6.0),   # Mg ha-1
  diameter_cm = c(22.0, 18.5, 15.0)
)

mySim <- simInit(
  times   = list(start = 0, end = 50),
  params  = list(DeadWood_snagDecay = list(species = "Pinus strobus")),
  modules = list("DeadWood_snagDecay"),
  objects = list(cohortData = cohortData)
)

mySim <- spades(mySim)
mySim$snagTable
mySim$fallenSnags
```

---

## References

Vanderwel, M.C., Malcolm, J.R., Smith, S.M., and Islam, N. (2006).
An integrated model for snag and downed woody
debris decay class transition.
*Forest Ecology and Management*, 234(1–3), 48–59.
<https://doi.org/10.1016/j.foreco.2006.06.020>

---

## Package dependencies

- [`SpaDES.core`](https://github.com/PredictiveEcology/SpaDES.core) (>= 3.0.0)
- [`data.table`](https://CRAN.R-project.org/package=data.table)
