---
title: "DeadWood_snagDecay"
author: ""
date: "`r Sys.Date()`"
output: github_document
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile, encoding = encoding, output_file = "README.md",
                      output_format = "github_document") })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Overview

`DeadWood_snagDecay` advances standing dead trees (snags) through five decay
classes (DC1–DC5) every 5 years using two independent stochastic processes.
First, each snag draws from a species-specific Markov transition matrix to
determine which decay class it moves to. Second, each snag independently draws
from a Bernoulli trial using a species- and DC-specific fall probability to
determine whether it falls that timestep. The two processes are independent: a
snag can advance decay classes and fall in the same timestep. Snags that fall
are transferred to `sim$fallenSnags` for ingestion by `DeadWood_DWDDecay`.

## Inputs

| Object | Class | Description |
|--------|-------|-------------|
| `cohortData` | `data.table` | Pixel-level tree mortality events. Required columns: `pixelID` (integer), `year` (numeric — year of death), `species` (character), `biomass` (numeric, Mg ha⁻¹ at death). Optional column: `diameter_cm` (numeric, cm; see `defaultDiameter_cm` parameter if absent). Typically produced by `DeadWood_Mortality`; can be supplied directly or replaced by any module that outputs `cohortData` (e.g. LandR Biomass). |

**Minimum diameter:** Any snag with `diameter_cm < 7.5 cm` will cause an error.
This threshold reflects the minimum piece size tracked in Vanderwel et al. (2006).

**Untracked species:** Rows in `cohortData` whose `species` value is not listed
in the `species` parameter are silently dropped each timestep. A runtime warning
will identify which species were ignored and prompt you to add them to the
`species` parameter if they should be tracked.

## Outputs

| Object | Class | Description |
|--------|-------|-------------|
| `snagTable` | `data.table` | Current standing snag inventory, updated every 5 years. |
| `fallenSnags` | `data.table` | Snags that fell during the current timestep, passed to `DeadWood_DWDDecay`. Reset each timestep. |

Both tables share the same schema:

| Column | Type | Description |
|--------|------|-------------|
| `pixelID` | integer | Raster pixel of origin |
| `species` | character | Tree species |
| `DC` | integer | Current decay class (1–5) |
| `ageInDC` | integer | Years spent in the current decay class |
| `initBiomass` | numeric | Biomass at time of death (Mg ha⁻¹); never modified after entry |
| `diameter_cm` | numeric | Piece diameter (cm); never modified after entry |

## Parameters

| Parameter | Type | Default | Range | Description |
|-----------|------|---------|-------|-------------|
| `snagTransMat` | matrix | See below | — | 5×5 Markov transition probability matrix for decay class advancement (5-year intervals). Source: Vanderwel et al. (2006) Table 3. |
| `snagFallProb` | numeric | See below | [0, 1] | Length-5 vector of 5-year fall probabilities, one per decay class. Source: Vanderwel et al. (2006) Table 2. |
| `species` | character | `c("Pinus strobus", "Pinus resinosa")` | — | Character vector of species to process from `cohortData`. Each entry must have a named entry in both `snagTransMat` and `snagFallProb`. |
| `defaultDiameter_cm` | numeric | `19.0` | ≥ 7.5 | Fallback diameter (cm) assigned to all incoming snags when `cohortData` lacks a `diameter_cm` column. |

### `snagTransMat`

A named list of 5×5 Markov transition matrices, one per species. Each matrix
gives the 5-year conditional probability of moving from one decay class (row)
to another (column). DC5 is absorbing — snags that reach DC5 remain there
until they fall. Row names are `from_DC1`–`from_DC5`; column names are
`to_DC1`–`to_DC5`. Names in the list must match the `species` parameter.

Transitions are strictly forward (lower triangle is zero); a snag cannot
revert to an earlier decay class.

Default matrices (Vanderwel et al. 2006, Table 3):

***Pinus strobus***

| | DC1 | DC2 | DC3 | DC4 | DC5 |
|---|---:|---:|---:|---:|---:|
| **from DC1** | 0.109 | 0.368 | 0.471 | 0.035 | 0.017 |
| **from DC2** | 0.000 | 0.348 | 0.601 | 0.051 | 0.000 |
| **from DC3** | 0.000 | 0.000 | 0.704 | 0.270 | 0.027 |
| **from DC4** | 0.000 | 0.000 | 0.000 | 0.913 | 0.087 |
| **from DC5** | 0.000 | 0.000 | 0.000 | 0.000 | 1.000 |

***Pinus resinosa***

| | DC1 | DC2 | DC3 | DC4 | DC5 |
|---|---:|---:|---:|---:|---:|
| **from DC1** | 0.109 | 0.368 | 0.471 | 0.035 | 0.017 |
| **from DC2** | 0.000 | 0.234 | 0.434 | 0.252 | 0.081 |
| **from DC3** | 0.000 | 0.000 | 0.704 | 0.270 | 0.027 |
| **from DC4** | 0.000 | 0.000 | 0.000 | 0.913 | 0.087 |
| **from DC5** | 0.000 | 0.000 | 0.000 | 0.000 | 1.000 |

The two species share identical DC1, DC3, DC4, and DC5 rows; they differ at
DC2, where *P. resinosa* transitions more rapidly to DC4–DC5.

### `snagFallProb`

A named list of 5-year fall probability vectors, one per species. Each vector
has length 5, with one probability per decay class. At each 5-year timestep
each snag is independently sampled using its current DC's fall probability;
snags that fall are removed from `snagTable` and added to `fallenSnags`.

Default values (Vanderwel et al. 2006, Table 2):

| Species | DC1 | DC2 | DC3 | DC4 | DC5 |
|---------|---:|---:|---:|---:|---:|
| *Pinus strobus* | 0.130 | 0.164 | 0.212 | 0.325 | 0.200 |
| *Pinus resinosa* | 0.000 | 0.039 | 0.094 | 0.224 | 0.080 |

*P. resinosa* DC1 has a fall probability of zero — freshly dead red pine snags
do not fall immediately. Fall probability generally increases with decay class,
peaking at DC4 for both species.

### `species`

Character vector of species names to process from `cohortData`. Default:
`c("Pinus strobus", "Pinus resinosa")`. Each entry must have a corresponding
named entry in both `snagTransMat` and `snagFallProb`; the module will stop
with an informative error if a name is missing.

### `defaultDiameter_cm`

Fallback piece diameter (cm) applied to all incoming snags when `cohortData`
does not include a `diameter_cm` column. Default: `19.0` cm. Minimum allowed
value is `7.5` cm (the minimum coarse woody debris threshold). Diameter is
passed to `DeadWood_DWDDecay` and controls the logistic transition rate once
a piece falls.

## Events

The module fires two event types on a 5-year schedule.

### `init` (once, at simulation start)

- Validates that each species in `species` has a named entry in `snagTransMat`
  and `snagFallProb`, that each transition matrix is 5×5 with no all-zero rows,
  and each fall probability vector has length 5 with values in [0, 1].
- Creates empty `snagTable` and `fallenSnags` with the correct schema.
- Schedules the first `transition` event at `start(sim) + 5` with priority 1.

### `transition` (every 5 years, priority 1)

Executes three steps in order:

**Step 1 — Absorb new mortality**

Rows in `cohortData` matching the target `species` whose `year` falls in the
interval `(time - 5, time]` are treated as trees that died during the preceding
5-year period. Each is added to `snagTable` as a DC1 snag with `ageInDC = 0`.

**Step 2 — Advance decay class**

Each snag's decay class is updated by drawing from the row of `snagTransMat`
corresponding to its current DC.

$$P(DC_{\text{new}} = j \mid DC_{\text{old}} = i) = \text{snagTransMat}[i,\; j]$$

where $i$ is the snag's current decay class (1–5) and $j$ is the decay class
being transitioned into (1–5). The equation gives the probability of moving
from class $i$ to class $j$ in a single 5-year timestep, read directly from
row $i$, column $j$ of the transition matrix. One value of $j$ is drawn for
each snag.

Because transition probabilities only move forward (upper-triangular matrix), DC
never decreases. Time in decay class is tracked:

$$\text{ageInDC}_\text{new} = \begin{cases} \text{ageInDC} + 5 & \text{if } \text{DC}_\text{new} = \text{DC}_\text{old} \\ 0 & \text{if } \text{DC}_\text{new} > \text{DC}_\text{old} \end{cases}$$

**Step 3 — Stochastic falling**

Using the **pre-transition** DC, each snag independently draws a fall outcome.

$$P(\text{snag falls} \mid DC_{\text{old}} = i) = \text{snagFallProb}[i]$$

where $i$ is the snag's current decay class (1–5) before the transition step,
and `snagFallProb[i]` is the species-specific 5-year probability of falling
for a snag in class $i$. Each snag is evaluated independently as a Bernoulli
trial — it either falls or it does not.

Snags that fall are moved to `fallenSnags`. Snags that do not fall remain in
`snagTable` with their updated DC and `ageInDC`.

## Event scheduling and module interactions

This module is designed to run alongside `DeadWood_Mortality`, `DeadWood_DWDDecay`,
and `DeadWood_Biomass`. `DeadWood_Mortality` fires every year (priority 0);
the remaining modules fire every 5 years. Priority determines order only among
events scheduled at the same simulation time.

| Timestep | Priority | Module | Event | Purpose |
|----------|----------|--------|-------|---------|
| every 1 yr | 0 | `DeadWood_Mortality` | `transition` | Apply baseline mortality, append to `cohortData` |
| every 5 yr | 1 | `DeadWood_snagDecay` | `transition` | Absorb `cohortData` (5-yr window), advance snag DC ← **this module** |
| every 5 yr | 2 | `DeadWood_DWDDecay` | `receive` | Accept `fallenSnags` into DWD pool |
| every 5 yr | 3 | `DeadWood_DWDDecay` | `transition` | Advance DWD DC |
| every 5 yr | 4 | `DeadWood_Biomass` | `transition` | Compute biomass rasters |
| every 5 yr | 5 | `DeadWood_Biomass` | `plot` | Capture biomass snapshots |

This ordering guarantees that `fallenSnags` is populated before `DeadWood_DWDDecay`
reads it, and that biomass is computed from the fully updated inventories.

## Example

```{r example}
library(SpaDES.project)
set.seed(42)

# Three mortality pulses at different years fill the full DC spectrum by year 50:
#   Year  1 cohort: 50 years of decay → concentrated in DC4-5
#   Year 15 cohort: 35 years          → mixed DC2-4
#   Year 30 cohort: 20 years          → mostly DC1-2
cohortData <- data.table::data.table(
  pixelID     = 1:60,
  year        = c(rep(1L, 20), rep(15L, 20), rep(30L, 20)),
  species     = rep(c(rep("Pinus strobus", 10), rep("Pinus resinosa", 10)), 3),
  biomass     = c(rep(25.0, 20), rep(20.0, 20), rep(15.0, 20)),
  diameter_cm = c(rep(32.0, 20), rep(26.0, 20), rep(20.0, 20))
)

out <- SpaDES.project::setupProject(
  paths      = list(modulePath = "modules"),
  modules    = "BosunForestEcology/DeadWood_snagDecay@main",
  times      = list(start = 0, end = 50),
  cohortData = cohortData
)

mySim <- SpaDES.core::simInitAndSpades2(out)

library(ggplot2)

dcSummary <- mySim$snagTable[, .(biomass = sum(initBiomass)), by = .(DC, species)][order(species, DC)]

ggplot(dcSummary, aes(x = DC, y = biomass, colour = species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = 1:5) +
  scale_colour_manual(values = c("Pinus strobus" = "darkgreen", "Pinus resinosa" = "saddlebrown")) +
  labs(x = "Decay class", y = "Total snag biomass (Mg/ha)",
       title = "Snag biomass by decay class at year 50", colour = NULL) +
  theme_bw()
```

## References

Vanderwel, M.C., Malcolm, J.R., Smith, S.M., and Islam, N. (2006).
An integrated model for snag and downed woody debris decay class transition.
*Forest Ecology and Management*, 234(1–3), 48–59.
<https://doi.org/10.1016/j.foreco.2006.06.020>

## Package dependencies

- [`SpaDES.core`](https://github.com/PredictiveEcology/SpaDES.core) (>= 3.0.0)
- [`data.table`](https://CRAN.R-project.org/package=data.table)
